### 2024-10-03 esop25/metalang.tex
checkpoint

Our first goal with a $\metaprot$ type theory is to define an
algorithmic system that is sound for both confidentiality and
integrity typings as defined in Sections \ref{section-cpj} and
\ref{section-ipj}. Returning to the example in Section
\ref{section-cpj}, note that the key equivalence of the $\ttt{mux}$
expression with a one-time-pad encryption we observe there
is not trivial. To ensure that this sort of equivalence
can be picked up by the type system, we introduce a
an annotation form that allows the programmer to provide
the needed hint.

Consider the following $\ttt{encode}$ function definition that
generalizes this encoding for any identifier $y$ and sender, receiver
pair $s,r$:
$$
\begin{array}{l}
\ttt{encode}(y,s,r) \{\\
\quad \xassign{\mx{y}{r}}{\mux{\secret{y}}{\neg\flip{y}}{\flip{y}}}{s};\\
\quad \eqcast{\mx{y}{r}}{\neg\sx{y}{s} \fplus \rx{y}{s}}\\
\}
\end{array}
$$

Furthermore, we want to minimize the amount of SMT solving needed for
type verification.  Our approach is to support pre- and post-condition
annotations on $\metaprot$ functions that can be independently
verified, allowing significant reduction of SMT overhead in
whole-program analysis.

@@

@@

\begin{lemma}
  If $\atj{\toeq{\phi}}{R}{\ty}$ then $\eqj{R}{\eqs}{\phi}{\ty}$ for any $\eqs$.
\end{lemma}

$$
\cmd ::= \cdots \mid \eqcast{\mx{e}{e}}{\notg{\phi}}
$$

>>>>

Our first goal in the $\metaprot$ type theory is to define an
algorithmic system that is sound for both confidentiality and
integrity typings as defined in Sections \ref{section-cpj} and
\ref{section-ipj}. Returning to the example in Section
\ref{section-cpj}, note that the key equivalence of the $\ttt{mux}$
expression with a one-time-pad encryption we observe there
is not trivial. To ensure that this sort of equivalence
can be picked up by the type system, we introduce a
an annotation form that allows the programmer to provide
the needed hint.

Consider the following $\ttt{encode}$ function definition that
generalizes this encoding for any identifier $y$ and sender, receiver
pair $s,r$. The second line in the body is a hint that
the message $\mx{y}{r}$ can be considered equivalent to
$\neg\sx{y}{s} \fplus \rx{y}{s}$:
$$
\begin{array}{l}
\ttt{encode}(y,s,r) \{\\
\quad \xassign{\mx{y}{r}}{\mux{\secret{y}}{\neg\flip{y}}{\flip{y}}}{s};\\
\quad \eqcast{\mx{y}{r}}{\neg\sx{y}{s} \fplus \rx{y}{s}}\\
\}
\end{array}
$$
This hint can be validated using SMT, and then the
syntactic structure of $\neg\sx{y}{s} \fplus \rx{y}{s}$
allows its immediate interpretation as a one-time-pad encoding.

Our second goal in the $\metaprot$ type system is to minimize the
amount of SMT solving needed for type verification.  Returning to the
$\ttt{encode}$ example, type checking in our systems only requires the
hint to be verified once, with the guarantee that $\ttt{encode}$ can
be applied anywhere without needing to re-verify the hint in
application instances. To verify the hint in $\ttt{encode}$, we can
just choose arbitrary values $w,\cid_1,\cid_2$ for $y$, $s$, and $r$,
evaluate $\ttt{encode}(w,\cid_1,\cid_2) \redx \prog$, and then verify:
$$\toeq{\prog} \models \mx{w}{\cid_2} \eop
\neg\sx{w}{\cid_1} \fplus \rx{w}{\cid_1}$$ 
Since $\ttt{encode}$ is closed, validity is guaranteed for
any instance of $y$, $s$, and $r$.

We generalize these benefits of compositional verification by allowing
pre- and post-condition annotations on $\metaprot$ functions. For
example, consider a GMW-style, 2-party ``and-gate'' function
$\ttt{andgmw}(x,y,z)$. In this protocol, each party $\cid
\in \{1,2\}$ holds an additive secret share $\mx{x}{\cid}$ of
values identified by $x$ and $y$, and at the end of the
protocol each hold a secret share $\mx{z}{\cid}$, where:
$$
\mx{z}{1} \fplus \mx{z}{2} \eop (\mx{x}{1} \fplus \mx{x}{2}) \ftimes (\mx{y}{1} \fplus \mx{y}{2})
$$
We provide details of $\ttt{andgmw}$ and other GMW protocol
elements in Section \ref{section-examples}.  Our point now is that,
similarly to the $\ttt{encode}$ example, we can verify this
post-condition once as a correctness property for $\ttt{andgmw}$, and
then integrate instances of it into circuit correctness properties
with the guarantee that each instance also holds for any
$\ttt{andgmw}$ gate.  Since the program logic of $\ttt{andgmw}$ is
non-trivial, and typical circuits can use up to thousands of gates,
this has significant practical benefits by greatly reducing SMT overhead
in whole-program analysis.

@@

A core element of $\metaprot$ type checking is type checking
of $\minicat$ expressions. The integrity type system presented in Section
\ref{section-ipj} is already algorithmic and ready to use. But confidentiality
typing presented in Section \ref{section-cpj} is not syntax-directed
due to the \TirName{Encode} rule. But as described above, by introducing
hint annotations we can ``cast'' any relevant expression form into
the simplest equivalent syntactic form for one-time-pad encoding.
Thus, in Figure \ref{fig-atj} we present the algorithm for
confidentiality type judgements, where we eliminate the need for
integrated SMT solving by assuming this sort of casting.

@@

The following Lemma establishes correctness of algorithmic confidentiality
type checking, and makes explicit that SMT checking is eliminated in the
judgement.
\begin{lemma}
  If $\atj{\toeq{\phi}}{R}{\ty}$ then $\eqj{R}{\eqs}{\phi}{\ty}$ for any $\eqs$.
\end{lemma}
